# Advanced Usage

This chapter shows advanced usage of `r pkg("normpatch")`,
including how to define custom normalization and test functions for `calc_nf()`
and how to pass additional parameters to the DEGES workflow.


```{r advcs-setup, include=FALSE}
source("_config.R")
```


## Custom normalization functions

`r pkg("normpatch")` provides DEGES-based normalization workflows.
The workflow requires both a normalization function and a DEG screening function.
By default, `r pkg("normpatch")` uses normalization and test functions from widely used packages,
`r pkg("edgeR", "bioc")` and `r pkg("DESeq2", "bioc")`.
In addition to the provided functions,
`r pkg("normpatch")` also supports custom functions.

Here, we show an example of using a custom normalization function with `calc_nf()`.
The normalization function should receive `x` as a count matrix
and `exp_design` as an experimental design data frame,
then return normalization factors.
Following this rule, users can define their own normalization steps.

Here, we implement upper-quartile normalization and pass it to `calc_nf()`.

```{r advcs-uq-func-def}
norm_uq <- function(x, exp_design, ...) {
    x <- as.matrix(x)
    keep <- rowSums(x > 0) > 0
    x <- x[keep, , drop = FALSE]
    uq <- apply(x, 2, quantile, probs = 0.75, names = FALSE)
    nf <- uq / colSums(x)
    nf <- nf / exp(mean(log(nf)))
    nf
}
```

In this example, we customize only the normalization function.
The test function uses the default `r pkg("edgeR", "bioc")` quasi-likelihood test,
so no custom test function is needed.

After defining the function, load the required packages and prepare the data as in the basic workflow.

```{r advcs-uq-framework, message=FALSE, warning=FALSE}
library(edgeR)
library(baySeq)
library(normpatch)

count_data <- read.table("data/ath.E-MTAB-4391.txt",
                         header = TRUE, sep = "\t", row.names = 1)
keep <- rowSums(edgeR::cpm(count_data) > 5) > (ncol(count_data) / 2)
count_data <- as.matrix(count_data[keep, , drop = FALSE])

group <- sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = factor(group))

x <- newSeqCountData(count_data, exp_design = exp_design)
```

Then run `calc_nf()` with the custom `norm_uq()` function specified
through the `norm_func` argument.


```{r advcs-uq-framework-calcnf, message=FALSE, warning=FALSE}
x <- calc_nf(x, norm_func = norm_uq)
x@meta$nf
```

The example above implements upper-quartile normalization
to demonstrate how a custom normalization function can be integrated into `calc_nf()`.

However, `r pkg("edgeR", "bioc")` can also perform upper-quartile normalization
through `calcNormFactors()` by setting `method = "upperquartile"`.
The workflow can therefore be changed from TMM to upper-quartile normalization as follows.
Additional arguments supplied to `calc_nf()` are forwarded to the normalization function.
In this example, `method = "upperquartile"` is passed to `calcNormFactors()`.

```{r advcs-uq-framework-edgeR, message=FALSE, warning=FALSE}
x <- newSeqCountData(count_data, exp_design = exp_design)
x <- calc_nf(x, method="upperquartile")
x@meta$nf
```


## Custom test functions

TbT [@TbT] is the original DEGES-based normalization workflow.
It estimates normalization factors using TMM and
screens candidate DEGs with `r pkg("baySeq", "bioc")`.
Because `r pkg("normpatch")` does not provide a built-in `r pkg("baySeq", "bioc")` screening function,
this section shows how to define a custom test function that reproduces the TbT workflow.

The test function should receive at least `x` as a count matrix,
`exp_design` as a data frame of experimental design,
and `nf` as normalization factors.
It should return at least p-value (`p_value`) and q-value (`q_value`) columns in a data frame.
Additional arguments can also be defined.
For example, `r pkg("baySeq", "bioc")` requires
`samplesize` for bootstrapping during the test
and cluster information through `cl` for parallel computation.
The custom function adds these arguments and sets their default values.


```{r advcs-test-bayseq-def, results='hide', message=FALSE, warning=FALSE}
library(baySeq)
library(parallel)

test_bayseq <- function(x, exp_design, nf, samplesize=10000, cl=NULL, ...) {
    d <- new("countData",
            data = x,
            replicates = exp_design$group,
            groups = list(NDE = rep(1, nrow(exp_design)), DE = exp_design$group),
            libsizes = nf * colSums(x))
    d@annotation <- data.frame(gene_id = 1:nrow(x))
    d <- getPriors.NB(d, estimation = "QL", samplesize = samplesize, cl = cl)
    d <- baySeq::getLikelihoods(d, pET = "BIC", cl = cl)
    test_stats <- topCounts(d, group = "DE", number = nrow(x))
    test_stats <- test_stats[order(test_stats$gene_id), ]
    p <- test_stats$FWER.DE
    q <- test_stats$FDR.DE
    p[is.na(p)] <- 1
    q[is.na(q)] <- 1
    data.frame(
        p_value = p,
        q_value = q,
        row.names = NULL, check.names = FALSE)
}
```

For the normalization step, the built-in `norm_tmm()` function can be used directly.
After defining the test function, load the data and run normalization.

```{r advcs-tbt-framework, results='hide', message=FALSE, warning=FALSE}
library(edgeR) # to use 'cpm' function
library(normpatch)

count_data <- read.table("data/ath.E-MTAB-4391.txt",
                         header = TRUE, sep = "\t", row.names = 1)
keep <- rowSums(cpm(count_data) > 5) > (ncol(count_data) / 2)
count_data <- as.matrix(count_data[keep, , drop = FALSE])

group <- sapply(strsplit(colnames(count_data), "_"), "[", 1)
exp_design <- data.frame(group = factor(group))

x <- newSeqCountData(count_data, exp_design = exp_design)

if(require("parallel")) {
    cl <- makeCluster(4)
} else {
    cl <- NULL
}

x <- calc_nf(x, norm_func = norm_tmm, test_func = test_bayseq,
             iter = 1, samplesize = 100, cl = cl)

if (!is.null(cl)) {
    stopCluster(cl)
}
```

Although `test_bayseq()` is primarily designed for use within `calc_nf()` during DEGES-based normalization,
it also returns gene-level p-values and q-values.
Therefore, it can also be used directly for DEG analysis.

```{r advcs-tbt-framework-detection, results='hide', message=FALSE, warning=FALSE}
if(require("parallel")) {
    cl <- makeCluster(4)
} else {
    cl <- NULL
}

test_stats <- test_bayseq(x@data, x@exp_design, x@meta$nf, samplesize = 100, cl = cl)

if (!is.null(cl)) {
    stopCluster(cl)
}

test_stats <- data.frame(gene_name = x@gene_names, test_stats)
```

```{r advcs-tbt-output}
head(test_stats)
```

Next, visualize the test results by highlighting genes with `q_value < 0.05`
in the `r pkg("baySeq", "bioc")` results.

```{r advcs-tbt-viz}
is_deg <- test_stats$q_value < 0.05
gene_colors <- ifelse(is_deg, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```

Here, `samplesize = 100` is used only for the example.
In practical analyses, use a larger value such as 10,000.


## GLM for multi-factor comparison

For experiments with multiple experimental factors,
the statistical model should be chosen to match the biological question.
You can supply different models to screen candidate DEGs for specific effects during normalization.

Here, we load and prepare a dataset for a multi-factor comparison.

```{r advcs-mfc-soybean-load-data}
atlas <- getAtlasData("E-GEOD-69469")
se <- atlas[[1]]$rnaseq

count_data <- as.matrix(assay(se, "counts"))
keep <- rowSums(cpm(count_data) > 5) > ncol(count_data) / 2
count_data <- count_data[keep, , drop = FALSE]
count_data <- as.matrix(count_data)

col_data <- as.data.frame(colData(se))
exp_design <- data.frame(
    group = factor(ifelse(grepl('control', col_data[, c("environmental_stress")]), 'control', 'stress')),
    timepoint = factor(gsub("Zeitgeber time ", "", col_data[, c("sampling_time_point")])))

x <- newSeqCountData(count_data, exp_design = exp_design)
x
```


By default, `calc_nf()` screens candidate DEGs using the available experimental design.
This can exclude genes associated with any modeled additive or interaction effects during DEGES-based normalization.

```{r cs-mfc-soybean-calcnf-default}
x <- calc_nf(x)
x@meta$nf
```

In some analyses, the normalization step should screen candidate DEGs for a specific factor only.
In that case, pass additional arguments to the test function used by `calc_nf()`.

For example, using TMM normalization and a `r pkg("edgeR", "bioc")` quasi-likelihood test,
the workflow below removes candidate DEGs for the treatment effect while adjusting for time point.

```{r cs-mfc-soybean-calcnf-edgeR-treatment-1}
design <- model.matrix(~ group + timepoint, data = exp_design)
head(design)
```

After confirming that the second column of the design matrix represents the treatment effect,
pass `design` and `coef = 2` to `calc_nf()`.

```{r cs-mfc-soybean-calcnf-edgeR-treatment-2}
x <- calc_nf(x, design = design, coef = 2)
x@meta$nf
```

See the `r pkg("edgeR", "bioc")` documentation for more detailed guidance on model matrices,
coefficients, and contrasts.

`r pkg("DESeq2", "bioc")` uses a different testing interface.
For an LRT, specify a full model and a reduced model.
The test evaluates whether terms present only in the full model explain additional variation.
For example, to test the treatment effect after adjusting for time point, use the following workflow.


```{r cs-mfc-soybean-calcnf-DESeq2}
x <- newSeqCountData(count_data, exp_design = exp_design)
x <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2, full = ~ group + timepoint, reduced = ~ timepoint)
x@meta$nf
```
