# Simulation Evaluation

This chapter introduces the simulation utilities in `r pkg("normpatch")`.
The two main functions are `sim_gene_counts()` and `def_DEG()`.
`sim_gene_counts()` generates artificial RNA-seq count matrices,
and `def_DEG()` defines the true DEG status from the simulation parameters and a fold change threshold.
These functions are useful for checking normalization behavior under controlled settings,
especially when DEGs are intentionally biased toward one group.

The examples in this chapter are intended as practical templates.
For a formal benchmark, repeat each simulation many times, vary the simulation parameters,
and summarize performance across simulation replicates rather than relying on one dataset.


## Generate simulated data

### Standard usage

The simplest call to `sim_gene_counts()` creates
a two-group count matrix containing 20,000 genes with three replicates per group.
The function returns a `SeqCountData` object whose `data` slot contains the count matrix
and whose `exp_design` slot contains the sample grouping.

```{r sim-std-tgc}
library(normpatch)
set.seed(1)

x <- sim_gene_counts()
x
```

The simulation parameters are stored in `x@meta$sim_params`.
They can be inspected through the slots of this object.


```{r sim-std-tgc-simparams}
x@meta$sim_params@n_genes
x@meta$sim_params@p_DEG
head(x@meta$sim_params@params)
head(x@meta$sim_params@fc)
```

Because the true mean and fold change parameters are known,
DEG status can be defined directly from the simulated fold change matrix.
The `def_DEG()` function marks genes as true DEGs when their largest between-group fold change
is at least the cutoff specified by `fc`.
`fc` defaults to 2 but can be changed.

```{r sim-std-tgc-degdef}
is_DEG <- def_DEG(x)
head(is_DEG)
```

An MA plot provides a direct view of the simulated counts and the known true DEGs.
Because `plot_ma()` uses normalized counts, estimate normalization factors before plotting.
Here, simulated DEGs are highlighted in red and genes that are not DEGs are shown in gray.

```{r sim-std-tgc-viz}
x <- calc_nf(x)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```


The number of genes and the degree of DEG imbalance can be changed by specifying the corresponding arguments.
In the following example, the second group contains a larger proportion of biased DEGs,
and those genes are assigned a stronger fold change.

```{r sim-std-tgc-args}
set.seed(1)

x <- sim_gene_counts(
    n_genes = 2000,
    n_replicates = c(4, 4),
    p_DEG = c(0.05, 0.15),
    fc_mean = c(8, 2),
    group_names = c("control", "stress")
)

is_DEG <- def_DEG(x, fc = 2)

x <- calc_nf(x)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```


To generate a comparison with three groups, provide one replicate count, DEG proportion,
and fold change value for each group.
The example below creates three groups with different proportions of DEGs.

```{r sim-std-mgc}
set.seed(1)

x <- sim_gene_counts(
    n_replicates = c(3, 3, 3),
    p_DEG = c(0.05, 0.10, 0.15),
    fc_mean = c(4, 4, 4),
    group_names = c("leaf", "root", "shoot")
)
x
```


### Generate counts with randomized fold changes

#### Log-normal fold changes

By default, `sim_gene_counts()` samples fold changes for genes assigned as DEGs from a lognormal distribution.
This makes the simulated effect sizes heterogeneous,
which is closer to typical RNA-seq datasets than assigning the same fold change to every DEG.

With randomized fold changes, `fc_mean` is interpreted as the mean fold change for genes assigned as DEGs.
For this distribution, the `fc_sd` argument controls the spread of the sampled values
as the standard deviation on the natural log fold change scale.


```{r sim-random-fc-lognormal}
set.seed(1)

x <- sim_gene_counts(
    fc_dist = "lognormal",
    fc_mean = c(4, 4),
    fc_sd = 0.5,
    group_names = c("control", "stress")
)

true_fc <- x@meta$sim_params@fc[, "stress"]
summary(true_fc[true_fc != 1])
```

The sampled fold-change matrix is stored in `x@meta$sim_params@fc`.
The `def_DEG()` function uses these fold changes to define the true DEG status.
Because this classification is based on a fold-change cutoff,
the resulting number of true DEGs may differ from the proportion specified by the `p_DEG` argument.


```{r sim-random-fc-viz}
is_DEG <- def_DEG(x, fc = 2)
table(is_DEG)

x <- calc_nf(x, iter = 0)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors) + ggtitle("lognormal")
```


#### Gamma fold changes

The gamma option provides another positive distribution for fold changes.
Here, the mean fold change is again set by `fc_mean`,
while `fc_sd` controls the coefficient of variation.

```{r sim-random-fc-gamma}
set.seed(1)

x <- sim_gene_counts(
    fc_dist = "gamma",
    fc_mean = c(4, 4),
    fc_sd = 0.5,
    group_names = c("control", "stress")
)

true_fc <- x@meta$sim_params@fc[, "stress"]
summary(true_fc[true_fc != 1])

x <- calc_nf(x, iter = 0)

is_DEG <- def_DEG(x, fc = 2)
gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors) + ggtitle("gamma")
```


#### Fixed fold changes

Fixed fold changes are still available for controlled simulations.
Set `fc_dist = "fixed"` when every DEG assigned
to a group should receive the exact value in `fc_mean`.

```{r sim-fixed-fc}
set.seed(1)

x <- sim_gene_counts(
    fc_dist = "fixed",
    fc_mean = c(4, 4),
    group_names = c("control", "stress")
)

true_fc <- x@meta$sim_params@fc[, "stress"]
unique(true_fc[true_fc != 1])

x <- calc_nf(x, iter = 0)

is_DEG <- def_DEG(x, fc = 2)
gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors) + ggtitle("fixed")
```

## Custom seed matrices

By default, `r pkg("normpatch")` uses the bundled `arab` dataset,
an *Arabidopsis thaliana* RNA-seq count matrix,
to construct the population of mean and variance parameters used for simulation.

```{r sim-load-arab}
data(arab, package = "normpatch")
head(arab)
```

The `sim_gene_counts()` function can also use a user-supplied count matrix
as the seed population for simulation.
The example below uses the *Arabidopsis* count matrix stored in `docs/data/ath.E-MTAB-4391.txt`
as the seed matrix.

```{r sim-load-seedmatrix}
seed_counts <- read.table("data/ath.E-MTAB-4391.txt", header = TRUE, sep = "\t", row.names = 1)
seed_counts <- as.matrix(seed_counts)
seed_counts <- rbind(seed_counts[, 1:3], seed_counts[, 4:6])
seed_counts <- seed_counts[rowSums(seed_counts) > 0, , drop = FALSE]
head(seed_counts)
```

The simulated count matrix can then be generated from this seed.

```{r sim-custom-data-generate}
set.seed(1)

x <- sim_gene_counts(seed_counts = seed_counts, fc_mean = c(2, 2))
is_DEG <- def_DEG(x, fc = 2)

x <- calc_nf(x)

gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")
plot_ma(x, col = gene_colors)
```


## Workflow evaluation

Because the true simulation parameters are known,
`def_DEG()` can be used to determine the reference DEG status of each gene.
This allows us to evaluate how well a DEG workflow ranks true DEGs ahead of non-DEGs.

In this section, we use simulated count data to evaluate DEG detection performance
using the area under the receiver operating characteristic curve, or AUC.
An AUC of 0.5 indicates performance similar to random ranking,
whereas values closer to 1 indicate that true DEGs are ranked more accurately.

First, generate a simulated count dataset.

```{r sim-eval-data}
set.seed(1)

x <- sim_gene_counts(n_genes = 20000, p_DEG = c(0.01, 0.04), fc_mean = c(4, 4))
is_DEG <- def_DEG(x, fc = 4)

table(is_DEG)
```

The following MA plot highlights the simulated true DEGs.

```{r sim-eval-data-viz, echo=FALSE, include=TRUE}
x_norm <- calc_nf(x)
gene_colors <- ifelse(is_DEG, "#B24745", "#ADADAD")

plot_ma(x_norm, col = gene_colors)
```

Next, define a helper function that calculates the AUC from gene-level p-values.
Smaller p-values are converted to larger ranking scores
so that genes with stronger statistical evidence are ranked first.

```{r sim-eval-auc-helper}
library(pROC)

calc_auc <- function(is_deg, p_value) {
    p_value[is.na(p_value)] <- 1
    ranking_score <- -p_value
    roc_obj <- roc(is_deg, ranking_score)
    auc_value <- auc(roc_obj)
    as.numeric(auc_value)
}
```


### Comparison with edgeR

This subsection uses simulated counts to compare a standard `r pkg("edgeR", "bioc")`
workflow with a DEGES-based normalization workflow followed by `r pkg("edgeR", "bioc")`
DEG testing.

First, run TMM normalization and a quasi-likelihood test with `r pkg("edgeR", "bioc")`,
then compute the AUC.

```{r sim-eval-edger-edger}
library(edgeR)

design <- model.matrix(~ group, data = x@exp_design)
d1 <- DGEList(counts = x@data, group = x@exp_design$group)
d1 <- calcNormFactors(d1)
d1 <- estimateDisp(d1, design)
fit1 <- glmQLFit(d1, design)
qlf1 <- glmQLFTest(fit1, coef = 2)
edger_output1 <- topTags(qlf1, n = nrow(d1$counts), sort.by = "none")$table

calc_auc(is_DEG, edger_output1$PValue)
```

Next, estimate normalization factors with the DEGES approach using `calc_nf()`.
The result is converted to a `r pkg("edgeR", "bioc")` object for the downstream DEG test,
and AUC is computed from the resulting p-values.

```{r sim-eval-edger-deges}
x <- calc_nf(x, orm_func = norm_tmm, test_func = test_edger)

design <- model.matrix(~ group, data = x@exp_design)
d2 <- as_DGEList(x)
d2 <- estimateDisp(d2, design)
fit2 <- glmQLFit(d2, design)
qlf2 <- glmQLFTest(fit2, coef = 2)
edger_output2 <- topTags(qlf2, n = nrow(d2$counts), sort.by = "none")$table

calc_auc(is_DEG, edger_output2$PValue)
```


### Comparison with DESeq2

`r pkg("DESeq2", "bioc")` workflows can be evaluated in the same way.
In `r pkg("normpatch")`, the built-in `test_deseq2()` function uses an LRT
for DEG screening during DEGES-based normalization.
Here, four workflows are compared:

1. RLE normalization, followed by a `r pkg("DESeq2", "bioc")` Wald test.
2. RLE normalization, followed by a `r pkg("DESeq2", "bioc")` LRT.
3. DEGES using RLE normalization and `r pkg("DESeq2", "bioc")` Wald screening, followed by a `r pkg("DESeq2", "bioc")` Wald test.
4. DEGES using RLE normalization and `r pkg("DESeq2", "bioc")` LRT screening, followed by a `r pkg("DESeq2", "bioc")` LRT.


First, run the `r pkg("DESeq2", "bioc")` Wald test and compute AUC.

```{r sim-eval-deseq2-Wald}
library(DESeq2)

col_data <- x@exp_design
rownames(col_data) <- colnames(x@data)
design = ~ group

dds1 <- DESeqDataSetFromMatrix(countData = x@data, colData = col_data, design = design)
dds1 <- estimateSizeFactors(dds1)
dds1 <- estimateDispersions(dds1)
dds1 <- nbinomWaldTest(dds1)
deseq2_output_1 = as.data.frame(results(dds1))

calc_auc(is_DEG, deseq2_output_1$pvalue)
```

Using the same input data, run `r pkg("DESeq2", "bioc")` with an LRT.

```{r sim-eval-deseq2-LRT}
dds2 <- DESeqDataSetFromMatrix(countData = x@data, colData = col_data, design = design)
dds2 <- estimateSizeFactors(dds2)
dds2 <- estimateDispersions(dds2)
dds2 <- nbinomLRT(dds2, full = ~ group, reduced = ~ 1)
deseq2_output_2 = as.data.frame(results(dds2))

calc_auc(is_DEG, deseq2_output_2$pvalue)
```

Next, evaluate a DEGES workflow that uses a Wald test during DEG screening.
Because `r pkg("normpatch")` does not provide a built-in Wald screening function for `calc_nf()`,
the example first defines `test_deseq2_wald()`.

```{r sim-eval-deseq2-wald-deges}
test_deseq2_wald <- function(x, exp_design, nf = NULL, ...) {
    if (is.null(nf)) nf <- rep(1, ncol(x))

    x <- as.matrix(x)
    exp_design <- as.data.frame(exp_design)
    rownames(exp_design) <- colnames(x)

    effective_lib_sizes <- nf * colSums(x)

    dds <- DESeqDataSetFromMatrix(
        countData = x,
        colData = exp_design,
        design = ~ group
    )
    sizeFactors(dds) <- effective_lib_sizes / mean(effective_lib_sizes)
    dds <- estimateDispersions(dds)
    dds <- nbinomWaldTest(dds)

    test_output <- as.data.frame(results(dds))
    p <- test_output$pvalue
    p[is.na(p)] <- 1
    q <- stats::p.adjust(p, method = "BH")

    data.frame(
        p_value = p,
        q_value = q,
        row.names = NULL,
        check.names = FALSE
    )
}


x3 <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2_wald)
dds3 <- as_DESeqDataSet(x3)
dds3 <- estimateDispersions(dds3)
dds3 <- nbinomWaldTest(dds3)
deseq2_output_3 = as.data.frame(results(dds3))

calc_auc(is_DEG, deseq2_output_3$pvalue)
```

Finally, evaluate the DEGES workflow that uses the `r pkg("DESeq2", "bioc")`
LRT for DEG screening.

```{r sim-eval-deseq2-lrt-deges}
x4 <- calc_nf(x, norm_func = norm_rle, test_func = test_deseq2)
dds4 <- as_DESeqDataSet(x4)
dds4 <- estimateDispersions(dds4)
dds4 <- nbinomLRT(dds4, full = ~ group, reduced = ~ 1)
deseq2_output_4 = as.data.frame(results(dds4))

calc_auc(is_DEG, deseq2_output_4$pvalue)
```

The AUC values can then be summarized across the four workflows.


```{r sim-eval-deseq2-output-summary}
data.frame(
    pipeline = c("RLE + Wald", "RLE + LRT", "DEGES(Wald) + Wald", "DEGES(LRT) + LRT"),
    auc = c(
        calc_auc(is_DEG, deseq2_output_1$pvalue),
        calc_auc(is_DEG, deseq2_output_2$pvalue),
        calc_auc(is_DEG, deseq2_output_3$pvalue),
        calc_auc(is_DEG, deseq2_output_4$pvalue)
    )
)
```
